# Geoms {#sec-geoms}

```{r}
#| echo: false
#| message: false
#| results: asis
source("warning.R")
```

```{r}
#| echo: false
#| message: false
#| warning: false

library(conflicted)
```

A __geom__ (short for geometric object) is a component that defines how data is visually represented in a plot. Geoms determine the type of visualization or the graphical shape that will be drawn.

"These geoms are the fundamental building blocks of **ggplot2** [@wickham2024a]. They are useful in their own right, but are also used to construct more complex geoms. Most of these geoms are associated with a named plot: when that geom is used by itself in a plot, that plot has a special name." [@wickham2016]

ggplot2 already has a [long list of geoms](https://ggplot2.tidyverse.org/reference/index.html#geoms). We won't be discussing those unless there is an extension package that is an improvement to the original. Primarily, this chapter focuses on the geoms that ggplot2 does not include.

## Area plots

### Horizon chart

### Ribbon plot (improved)

### Streamgraph

## Bar charts

### Likert plot

### Mosaic plot

## Density plots

### Raincloud plot

### Ridgeline plot

## Geometric shapes

### Arc

### B-spline curve

### Bézier curve

### Circle

### Ellipse

## Heatmaps

### Calendar heatmap

## Intersection plots

### Upset plot

### Venn diagram

## Line charts

### Bump chart

### Dumbbell and lollipop plot

### Line chart with neon glow or shadow effects

### Slope chart

## Multivariate graphs

### Parallel coordinates plot (improved)

### Radar chart

## Pie charts

### Pie chart (improved)

### Moon chart

### Square chart / waffle chart

## Scatter plots

### Beeswarm plot

### Point density plot

## Text visualization

### Page layout visualization

### Word cloud

## Tree-based graphs

### Dendrogram

### Treemap

### Voronoi diagram
