# Geoms {#sec-geoms}

```{r}
#| echo: false
#| message: false
#| results: asis
source("warning.R")
```

```{r}
#| echo: false
#| message: false
#| warning: false

library(conflicted)
```

A **geom** (short for geometric object) is a component that defines how data is visually represented in a plot. Geoms determine the type of visualization or the graphical shape that will be drawn.

"These geoms are the fundamental building blocks of **ggplot2** [@wickham2024a]. They are useful in their own right, but are also used to construct more complex geoms. Most of these geoms are associated with a named [chart]: when that geom is used by itself in a [chart], that [chart] has a special name." [@wickham2016]

ggplot2 already has a [long list of geoms](https://ggplot2.tidyverse.org/reference/index.html#geoms). We won't be discussing those unless there is an extension package that is an improvement to the original. Primarily, this chapter focuses on the geoms that ggplot2 does not include.

## Area charts {#sec-area-charts}

In general, area charts are line charts

### Horizon chart {#sec-horizon-chart}

### Ribbon chart (improved) {#sec-ribbon-chart-improved}

### Streamgraph {#sec-streamgraph}

## Bar charts {#sec-bar-charts}

### Likert chart {#sec-likert-chart}

### Mosaic chart {#sec-mosaic-chart}

## Density charts {#sec-density-charts}

### Raincloud chart {#sec-raincloud-chart}

### Ridgeline chart {#sec-ridgeline-chart}

## Geometric shapes {#sec-geometric-shapes}

### Arc {#sec-arc}

### B-spline curve {#sec-b-spline-curve}

### Bézier curve {#sec-bezier-curve}

### Circle {#sec-circle}

### Ellipse {#sec-ellipse}

## Heatmaps {#sec-heatmaps}

### Calendar heatmap {#sec-calendar-heatmap}

## Intersection diagrams {#sec-intersection-diagrams}

### UpSet diagram {#sec-upset-diagram}

### Venn diagram {#sec-venn-diagram}

## Line charts {#sec-geoms}

### Bump chart {#sec-bump-chart}

### Dumbbell and lollipop chart {#sec-dumbbell-and-lollipop-chart}

### Line chart (with neon glow or shadow effects) {#sec-line-chart-with-neon-glow-or-shadow-effects}

### Slope chart {#sec-slope-chart}

## Multivariate charts {#sec-multivariate-charts}

### Parallel coordinates chart (improved) {#sec-parallel-coordinates-chart-improved}

### Radar chart {#sec-radar-chart}

## Pie charts {#sec-pie-charts}

### Pie chart (improved) {#sec-pie-chart-improved}

### Moon chart {#sec-moon-chart}

### Square chart (waffle chart) {#sec-square-chart-waffle-chart}

## Scatter charts {#sec-scatter-charts}

### Beeswarm chart {#sec-beeswarm-chart}

### Point density chart {#sec-point-density-chart}

## Text visualization {#sec-text-visualization}

### Page layout visualization {#sec-page-layout-visualization}

### Word cloud diagram {#sec-word-cloud-diagram}

## Tree diagrams {#sec-tree-diagrams}

### Dendrogram {#sec-dendrogram}

### Treemap diagram {#sec-treemap-diagram}

### Voronoi diagram {#sec-voronoi-diagram}
